# Schedule

This is the schedule for the course. 
Please note that it is liable to change (possibly at short notice). 
If you find a mismatch between this schedule and the official one^[https://mitsdu.sdu.dk/skema/activity/N100007101/e20], then it is the official one that is correct.

The *Textbook* column tells you the chapter(s) of the two textbooks that are most relevant to the session. The two textbooks are Gotelli (G)^[
Gotelli, NJ (2008) *A Primer of Ecology*. Fourth Edition, Sinauer Associates. ISBN: 978-0878933181] and Stearns and Hoekstra (SH)^[Stearns, SC & Hoekstra, RF (2005) *Evolution: An Introduction*. 2nd Edition, Oxford University Press. ISBN: 978-0199255634], and the numbers refer to the chapter numbers (e.g. SH.3, means Chapter 3 in Stearns and Hoekstra).

You should aim to read these chapters as the course proceeds.

```{r, echo = FALSE, include=knitr::is_latex_output()}
plot(1,1,type = "n",axes = FALSE,xlab="",ylab="")
text(1,1,labels = "The schedule is only available on the HTML version of this document")
```


```{r, echo = FALSE, message=FALSE,warning=FALSE,eval=knitr::is_html_output()}
library(flextable)
library(readxl)

system("cp -r ~/Dropbox/_SDU_Teaching/BB512/'Course Documents'/BB512_Schedule.xlsx ~/Dropbox/_SDU_Teaching/BB512/BB512_Book/BB512_Schedule.xlsx")
schedule <- readxl::read_excel("BB512_Schedule.xlsx")
```

```{r echo = FALSE}
library(lubridate)
library(calendar)
library(magrittr)
library(dplyr)
curl::curl_download("https://mitsdu.sdu.dk/skema/icalendar/activities/N100007101/e20/e20",destfile = "bb512.ics")

#Last sunday in October, 0300, CEST -> CET, 
#So the time offset moves from +2 to +1

savingsTimeSwitch <- with_tz(lubridate::as_datetime("2020-10-25 03:00:00"),"Europe/Copenhagen")

courseCal <- ic_read("bb512.ics") %>% 
  arrange(DTSTART) %>% 
  mutate(Session = row_number()) %>% 
  mutate(savingsT = DTSTART < savingsTimeSwitch) %>% 
  mutate(offsetT = if_else(savingsT == TRUE, 2, 1)) %>% #Check this every year!
  mutate(DTSTART2 = DTSTART + hours(offsetT)) %>% 
  mutate(DTEND2 = DTEND + hours(offsetT)) %>% 
  mutate(St = hour(DTSTART2)) %>% 
  mutate(En = hour(DTEND2)) %>% 
  mutate(wd = lubridate::wday(DTSTART2,label= TRUE)) %>% 
  mutate(Room = gsub(pattern = "Odense ",replacement = "", x = LOCATION)) %>% 
  mutate(Room = paste0("(",Room,")")) %>% 
  mutate(Date = paste0( month(DTSTART2,label = TRUE)," ",day(DTSTART2),", ",wd)) %>% 
  mutate(Time = paste0("kl.",St,"-",En)) %>% 
  select(Session,Date, Time, Room)

schedule <- readxl::read_excel("BB512_Schedule.xlsx")

schedule <- left_join(courseCal,schedule)
```



```{r, echo = FALSE, message=FALSE,eval=knitr::is_html_output()}
library(lubridate)
library(officer)

schedule2 <- schedule %>% 
  select(Part, Date, Time, Room, Session, Type, Topic, Instructor, Textbook)

big_b <- fp_border(color="gray70", width = 3)
std_b <- fp_border(color="gray70", style = "dashed")
big_b_end <- fp_border(color="white", width = 3)

ft <- flextable(schedule2, cwidth = c(2,9,4,1,.5,1,4,.75,.75)) %>% 
  bold(part = "header") %>% 
  bold(j = 1) %>% 
  color(i = ~ Type == "Lecture",
         j = ~  Session  + Type + Topic,
        color="red") %>% 
  merge_v(j = ~ Date + Part ) %>% 
  fontsize(part = "all", size = 14) %>% 
  vline(border = std_b, j = 1:4) %>% 
  rotate(j = ~ Part, align = 'top', rotation = 'tblr') %>% 
  hline(border = std_b ,i = seq(2,nrow(schedule2),2)) %>% 
  hline(border = big_b_end ,i =nrow(schedule2))

  
ft
```

```{r, echo =FALSE,message=FALSE,warning=FALSE}
source("personalCalendar.R")
```

