# Pre- and Post-reproduction census

## Background

Population growth estimates depend on when the census is taken relative to reproduction. A pre-breeding census counts individuals before reproduction; a post-breeding census counts them after reproduction. This changes how survival and fecundity are represented in the projection matrix, even if the underlying biology is the same.

Provided the conversion between the two census types is done correctly, the population growth rate ($\lambda$) they produce is identical — census timing is just a bookkeeping convention, not a biological difference. However, it is easy to get the conversion wrong, and doing so *does* change your estimate of $\lambda$. This chapter shows both: a correct conversion (same $\lambda$), and a common mistake (different $\lambda$).

## Learning outcomes

::: {.do-something}
Learning outcomes:

- Explain the difference between pre-breeding and post-breeding census models.
- Convert a post-breeding matrix to a pre-breeding matrix.
- Recognise that a correct conversion leaves $\lambda$ unchanged, and explain why.
- Identify how a mis-specified survival or fecundity term changes the resulting $\lambda$ estimate.
:::

## Worked example

### Inputs

We start with a **post-breeding census** matrix. In a post-breeding census, individuals are counted right after breeding, so the fecundity term for a stage is simply its fecundity scaled by *that stage's own* survival through the year (it has to survive the year in order to be alive, and to breed, at the moment of census).

```{r prepostbreeding, echo = TRUE}
p0 <- 0.2
p1 <- 0.9
p2 <- 0.6
m2 <- 3.0
m3 <- 6.0

A1 <- matrix(c(0.0, m2 * p1, m3 * p2,
               p0, 0.0, 0.0,
               0.0,p1,0), byrow = TRUE, nrow = 3)

```

### Steps

1. Define survival and fecundity values.
2. Build the projection matrix for one annual time step.
3. Use the matrix for projection and comparison tasks below.

### Output and interpretation

The resulting matrix encodes the post-breeding census assumptions for transitions and fecundity over one year.

Now let's correctly convert this to a **pre-breeding census** matrix. In a pre-breeding census, individuals are counted just *before* breeding, so two things change:

- Survival transitions shift by one stage, because an individual now has to survive through the year *before* being counted in the next stage at the census.
- Crucially, the fecundity term is no longer scaled by the *parent's* survival (the parent doesn't need to survive to be counted — the census happens before it would breed again). Instead, it is scaled by $p_0$, the survival of the *newborn offspring* through to the pre-breeding census.

```{r postpreconversion, echo = TRUE}
A2 <- matrix(c(0.0, m2 * p0, m3 * p0,
               p1, 0.0, 0.0,
               0.0,p2,0), byrow = TRUE, nrow = 3)
```

Compare population growth rates. This uses the `eigs` function from the `popdemo` R package.

```{r comparepgr, echo = TRUE}
(popdemo::eigs(A1, what = "lambda"))
(popdemo::eigs(A2, what = "lambda"))
```

Notice that $\lambda$ is (to rounding error) identical for `A1` and `A2`. This is not a coincidence: when the conversion between census types is done consistently, it only changes *how* the same underlying survival and reproduction are represented in the matrix — it does not change the biology, so it cannot change the population's actual growth rate.

## A common mistake

It is easy to get the pre-breeding conversion partly right and partly wrong. A very common error is to correctly shift the *survival* transitions to the next stage, but forget to also change the fecundity term — i.e. to keep scaling fecundity by the parent's own survival ($p_1$, $p_2$) instead of switching to the offspring survival ($p_0$):

```{r commonmistake, echo = TRUE}
A3 <- matrix(c(0.0, m2 * p1, m3 * p2,
               p1, 0.0, 0.0,
               0.0,p2,0), byrow = TRUE, nrow = 3)

(popdemo::eigs(A3, what = "lambda"))
```

This looks like a small, easy-to-miss slip — only the fecundity row is "wrong", and the matrix still looks like a sensible pre-breeding matrix. But it produces a substantially different (and incorrect) estimate of $\lambda$, compared to the correctly-converted `A2`.

## Your task

- Confirm that `A1` and `A2` give (to rounding error) the same $\lambda$. Explain, in your own words, why a *correct* change of census timing should not change the underlying population growth rate.
- Compare `A2` (correct pre-breeding matrix) and `A3` (the common mistake). Which specific matrix entries differ, and why does that particular mistake push $\lambda$ so far from the correct value?
- Modify one survival probability (e.g., `p1`) and re-build `A1`, `A2`, and `A3`. Does the correct conversion (`A1` vs `A2`) still agree? Does the size of the error from the mistake (`A2` vs `A3`) grow, shrink, or stay about the same?

## Takeaways

- Census timing (pre- vs. post-breeding) changes how fecundity and survival are encoded in the matrix, but a *correctly done* conversion between the two gives the same $\lambda$ — timing alone is a bookkeeping choice, not a biological difference.
- However, small errors in how survival or fecundity terms are calculated for a given census type are easy to make and can substantially bias your $\lambda$ estimate.
- The general lesson extends beyond census timing: the specific method used to calculate stage-specific survival and reproduction values for a matrix model matters. Applying the wrong calculation will bias your estimates, even if the matrix "looks" reasonable.
