# Continuous-time Lotka-Volterra Predator-Prey Model in R


## Background

The Lotka-Volterra model describes the interaction between predators and prey. This document demonstrates how to program the continuous-time version of the model in R. Then, by varying the parameters you can explore its features, and visualise the dynamics of predator and prey populations over time.


```{block, type="do-something"}
Learning outcomes: 

- Use of R to investigate an ecological model.
- Understanding how the Lotka-Volterra predator-prey model works, and how it is visualised.

```
## Your task

In R/RStudio, you will create an R script that programs the continuous time version of the Lotka-Volterra model. You will then use this model to investigate how the model works and how varying the parameters affects the system.

### Step 0: Create a new script

* Open R Studio and create a new script (File > New File > R Script). 
* Save it with a useful name (File > Save).
* Type or copy/paste the following code, step-by-step, into the Script.
* Run the code, or parts of it, selecting the parts you want to run and clicking "Run".

### Step 1: Load Necessary Libraries

We use the `deSolve` package to solve the system of differential equations numerically. Note that you may need to install the package first using the command `install.packages("deSolve")`.

```{r setup, include=TRUE}
library(deSolve)
```

### Step 2: Define the Model

The system of equations is defined as follows:

- Victim (prey) population growth: \( \frac{dV}{dt} = RV - aVC \)
- Consumer (predator) population growth: \( \frac{dC}{dt} = afVC - qC \)

The function below implements these equations in R.

```{r define-model}
predator_prey_model <- function(t, state, parameters) {
  with(as.list(c(state, parameters)), {
    dV <- R * V - a * V * C  # Victim (prey) growth equation
    dC <- a * f * V * C - q * C  # Consumer (predator) growth equation
    
    list(c(dV, dC))
  })
}
```

### Step 3: Set Parameters and Initial Conditions

We set the parameters and initial conditions for the model. I use the same nomenclature as the PDF from the [Excel exercise](https://jonesor.github.io/BB512_Book/lotka-volterra-predator-prey-dynamics.html). The parameters are as follows: *R* is the victim population growth rate, *a* is the attack rate of the predator, *f* is the conversion efficiency of the predator and *q* is the predator starvation (death) rate. For more detail about these, read the relevant section in the Neal textbook.

```{r parameters}
parameters <- c(R = 0.8,  # Victim (prey) growth rate
                a = 0.01, # Attack rate
                f = 0.008, # Conversion efficiency
                q = 0.1)  # Predator death rate

initial_state <- c(V = 1000,  # Initial victim population
                   C = 20)   # Initial consumer population
```

### Step 4: Simulate the Model

We simulate the model over a specified time period and store the results.

```{r simulation}
time <- seq(0, 100, by = 0.1)  # Time sequence
output <- ode(y = initial_state, times = time, func = predator_prey_model, parms = parameters)
output_df <- as.data.frame(output)
```

### Step 5: Visualise the Dynamics

How does the size of the two populations change through time? You should see that the **amplitude** and **period** change as the parameters change.

- **Amplitude**: Half the difference between a population's peak and trough. It shows how much the population fluctuates.
- **Period**: The time it takes for a population to complete one full cycle of oscillation. i.e. the peak-to-peak distance. It indicates the speed of the population cycles.

#### Population Dynamics Over Time

The plot below shows the changes in consumer and victim populations over time.

```{r population-dynamics}
# Plot victim population on the primary y-axis
plot(output_df$time, output_df$V, type = "l", col = "blue", lwd = 2, 
     ylab = "Victim Population (V)", xlab = "Time", main = "Predator-Prey Dynamics")

# Add consumer population on a secondary y-axis
par(new = TRUE)
plot(output_df$time, output_df$C, type = "l", col = "red", lwd = 2, 
     axes = FALSE, xlab = "", ylab = "")
axis(side = 4, col = "red", col.axis = "red")
mtext("Consumer Population (C)", side = 4, line = 3, col = "red")

# Add legend
legend("topright", legend = c("Victim (V)", "Consumer (C)"), 
       col = c("blue", "red"), lwd = 2)
```

#### Phase Plot with ZNGIs

The phase plot shows the relationship between victim and consumer populations. Zero Net Growth Isoclines (ZNGIs) indicate where the growth rate of each population is zero.

```{r phase-plot}
plot(output_df$V, output_df$C, type = "l", col = "purple", lwd = 2, 
     xlab = "Victim (V)", ylab = "Consumer (C)", 
     main = "Predator-Prey Dynamics with ZNGIs")

# Victim ZNGI
abline(v = parameters["q"] / (parameters["a"] * parameters["f"]), col = "blue", 
       lwd = 2, lty = 2)  
# Consumer ZNGI
abline(h = parameters["R"] / parameters["a"], col = "red", lwd = 2, lty = 2)  

# Add point for initial population sizes
points(output_df$V[1], output_df$C[1], pch = 19, col = "black", cex = 1.5)

# Add an arrow to the end of the line
arrows(x0 = output_df$V[nrow(output_df) - 1], 
       y0 = output_df$C[nrow(output_df) - 1], 
       x1 = output_df$V[nrow(output_df)], 
       y1 = output_df$C[nrow(output_df)], 
       col = "purple", length = 0.1, lwd = 2)
```

### Questions

**1. Oscillatory Dynamics**
- Run the simulation with the default parameters. Describe the oscillatory behavior of predator and prey populations over time. How do the populations interact?
- What happens to the *amplitude* and *period* of the oscillations if you double the initial prey population (\( V \))? What about doubling the initial predator population (\( C \))?

**2. Neutral Stability**
- Change the predator's death rate (\( q \)) and observe the long-term dynamics. Do the oscillations persist, amplify, or diminish? Why?

**3. Equilibrium Points**
- Identify the equilibrium point (coexistence) for the default parameters. Verify it by calculating the equilibrium populations using:
  \[
  V^* = \frac{q}{a \cdot f}, \quad C^* = \frac{R}{a}.
  \]
  Do the simulated populations approach these values?
- How do the equilibrium populations change when you increase the prey growth rate (\( R \)) or predator attack rate (\( a \))?

**4. Parameter Sensitivity**
- Investigate the effect of each parameter on the dynamics:
  - Alter \( R \) (prey growth rate). How does this affect the predator population?
  - Alter \( q \) (predator death rate). What happens to the predator-prey oscillations?
  - Alter \( f \) (conversion efficiency). What happens to the predator-prey oscillations?
  - Alter \( a \) (attack rate). What happens to the predator-prey oscillations?


**5. Phase Plane Analysis**
- Plot the phase plane with the default parameters. Add the Zero Net Growth Isoclines (ZNGIs) for predators and prey. What do these isoclines represent?
- Change the prey growth rate (\( R \)),  predator attack rate (\( a \)) and other parameters. How do the positions of the isoclines change? What about the shape and size of the orbits? What does this imply for the dynamics?
- Change the initial population sizes. Does the initial population size impact the dynamics?

**6. Extinction Scenarios**
- Set a very high predator death rate (\( q \)) or a very low attack rate (\( a \)). What happens to the predator population? Is it possible to drive the predators extinct? What are the implications for prey dynamics?
- What conditions (parameter combinations) are most likely to lead to extinction of both predators and prey?

**7. Real-World Relevance**
- Reflect on the assumptions of the model. What real-world factors might disrupt the neutrally stable oscillations predicted by the Lotka-Volterra model?



### Conclusion

This document demonstrates how to implement and explore the Lotka-Volterra predator-prey model in continuous time using R. By varying parameters and initial conditions, you can investigate how predator-prey dynamics are influenced by ecological factors. Notice that, unlike the discrete time model, which has an implicit time lag of 1, the continuous-time model results in dynamically stable cycles rather than cycles with increasing amplitude (and certain extinction). To fit a discrete-time version of the model, take a look at the Excel exercise ([here](https://jonesor.github.io/BB512_Book/lotka-volterra-predator-prey-dynamics.html)) or the R exercise (here).

The main findings of the continuous time model are as follows: 

1. **Oscillatory Dynamics**: Predator and prey populations exhibit cyclical oscillations. The prey population increases when predator numbers are low, while the predator population grows when prey is abundant.
   
2. **Neutral Stability**: The oscillations are neutrally stable under idealized conditions; they persist indefinitely without damping or amplification unless external factors or parameter changes are introduced. Unless the parameters set to truly extreme values, extinction is not possible.

3. **Equilibrium Points**: There are two equilibria:
   - The trivial equilibrium, where both predator and prey populations are zero.
   - The coexistence equilibrium, where the predator and prey populations remain constant over time.

4. **Parameter Dependence**: The amplitude and period of the oscillations depend on the parameters \( R \), \( a \), \( f \), and \( q \).

5. **Phase Plane Behavior**: Predator and prey populations follow closed orbits in the phase plane around the coexistence equilibrium, under the assumption that the system remains undisturbed.
