## Results of the hawk-dove games

I have crunched the numbers for 2022 and will summarise the results of the hawk-dove game.

If you are interested in playing with the data yourself you can find the data [here](https://www.dropbox.com/s/7ompsbe4iwirqar/gametheoryResults.xlsx?dl=0).



### Game 1

In this game, you were asked to compete with one person 15 times in a row.


```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(readxl)
library(janitor)

hdg <- readxl::read_xlsx("/Users/jones/Dropbox/_SDU_Teaching/BB512/Practicals/HawkAndDoveGame/gametheoryResults.xlsx", sheet = 3) %>%
  clean_names() %>%
  select(individual, round, player_1, player_2, benefit)


# Does hawkishness change during the game?
hawkishness <- hdg %>%
  mutate(hawkNumeric_player1 = ifelse(player_1 == "H", 1, 0)) %>%
  group_by(round) %>%
  summarise(meanHawk = mean(hawkNumeric_player1))


```
```{r echo=FALSE,results='hide'}
mod1 <- lm(meanHawk ~ round, data = hawkishness)
summary(mod1)
round(summary(mod1)$coefficients[2,4],3)
```

The average hawkishness, overall is `r round(mean(hawkishness$meanHawk),3)`.

We can now ask whether the level of hawkishness changed during the game on average across all the participants.

We can see from this graph that as the game proceeds round-by-round, there is no significant change in the level of hawkishness (p=`r round(summary(mod1)$coefficients[2,4],3)`).

```{r echo = FALSE}
ggplot(hawkishness,aes(x = round,y = meanHawk)) + 
  geom_point() + 
  geom_smooth(method = "lm") 
```
Next, we can ask whether the amount of benefits received (fitness) is associated with the degree of hawkishness.

What we see is that the maximum benefits I received when hawkishness is it an intermediate level of around 0.5-0.6.


```{r echo = FALSE}
# How does the hawkishness of individuals relate to the benefits recieved?
perIndiv <- hdg %>%
  mutate(hawkNumeric_player1 = ifelse(player_1 == "H", 1, 0)) %>%
  group_by(individual) %>%
  summarise(meanHawkishness = mean(hawkNumeric_player1), benefit = sum(benefit, na.rm = TRUE))
```

```{r echo = FALSE}
ggplot(perIndiv,aes(x = meanHawkishness,y = benefit)) + 
  geom_point() + 
  geom_smooth(method="lm", formula=y~poly(x, 2))
```

### GAME 2


In the next game you're asked to compete against a new opponent every round. The idea was that this would make it much harder to learn your opponents strategy and may be harder to come to a agreement that reduces aggression.


```{r echo = FALSE,warning=FALSE}
hdg2 <- readxl::read_xlsx("/Users/jones/Dropbox/_SDU_Teaching/BB512/Practicals/HawkAndDoveGame/gametheoryResults.xlsx", sheet = 4) %>%
  clean_names() %>%
  select(individual, round, player_1, player_2, benefit)
```




```{r echo = FALSE, results='hide'}
# Does hawkishness change during the game?
hawkishness <- hdg2 %>%
  mutate(hawkNumeric_player1 = ifelse(player_1 == "H", 1, 0)) %>%
  group_by(round) %>%
  summarise(meanHawk = mean(hawkNumeric_player1))

mod1 <- lm(meanHawk ~ round, data = hawkishness)
summary(mod1)
round(summary(mod1)$coefficients[2,4],3)


```

Firstly I have calculated the average hawkishness as  `r round(mean(hawkishness$meanHawk),3)`, which is not so different from the previous game.

Again, we can ask whether hawkishness changed during the game.

In this case, it DOES look like hawkishness declines more during the game. However, this trend is not significantly different from horizontal (p = `r round(summary(mod1)$coefficients[2,4],3)`).

```{r echo = FALSE}
ggplot(hawkishness,aes(x = round,y = meanHawk)) + 
  geom_point() + 
  geom_smooth(method = "lm")
```


Finally we can look at what the best strategy is in Game 2.



```{r echo = FALSE}
# How does the hawkishness of individuals relate to the benefits recieved?
perIndiv <- hdg2 %>%
  mutate(hawkNumeric_player1 = ifelse(player_1 == "H", 1, 0)) %>%
  group_by(individual) %>%
  summarise(meanHawkishness = mean(hawkNumeric_player1), benefit = sum(benefit, na.rm = TRUE))

ggplot(perIndiv,aes(x = meanHawkishness,y = benefit)) + 
  geom_point() + 
  geom_smooth(method="lm", formula=y~poly(x, 2))

```

It looks like the best strategy has changed a lot! Now the best strategy is to be a hawk.

In this setting we might expect  natural selection to drive towards the evolution of more aggressive individuals.





