## Results of the hawk-dove games

I have crunched the numbers for `r format(Sys.Date(), "%Y")` and will summarise the results of the hawk-dove game.


In the game, you were asked to compete with the same person 15 times in a row.


```{r hawkdoveresults1, echo = FALSE, message = FALSE}
library(tidyverse)
library(readxl)
library(janitor)

downloadNewData <- FALSE

if (downloadNewData == TRUE) {
  if (!requireNamespace("googledrive", quietly = TRUE)) {
    stop("Package 'googledrive' is required when downloadNewData is TRUE.")
  }
  # Download from Google Drive
  file <- googledrive::drive_find(pattern = "Game theory part 1", type = "spreadsheet")
  file
  googledrive::drive_download(
    file = googledrive::as_id(file$id),
    path = paste("data/", file$name, ".xlsx", sep = ""),
    type = "xlsx", overwrite = TRUE
  )
}

hdg <- readxl::read_xlsx("data/Game theory part 1 (Responses).xlsx", sheet = 1) %>%
  clean_names() 

names(hdg) <- c("timestamp","teamID","round","player_1_card","player_2_card","player_1_benefit", "player_2_benefit")

hdg <- hdg %>% 
  select(teamID, round, player_1_card, player_2_card, player_1_benefit, player_2_benefit)

#Filter name errors
fewRecordsTeams <- names(which(table(hdg$teamID)<10))

hdg <- hdg %>% 
  filter(!teamID %in% fewRecordsTeams)

# Does hawkishness change during the game?
hawkishness1 <- hdg %>%
  mutate(hawkNumeric_player1 = ifelse(player_1_card == "Hawk", 1, 0)) %>%
  group_by(round) %>%
  summarise(meanHawk = mean(hawkNumeric_player1))

hawkishness2 <- hdg %>%
  mutate(hawkNumeric_player2 = ifelse(player_2_card == "Hawk", 1, 0)) %>%
  group_by(round) %>%
  summarise(meanHawk = mean(hawkNumeric_player2))

hawkishness <- bind_rows(hawkishness1,hawkishness2)

# How does the average hawkishness of individuals relate to the benefits received?
perIndiv_player1 <- hdg %>%
  mutate(hawkNumeric_player1 = ifelse(player_1_card == "Hawk", 1, 0)) %>%
  group_by(teamID) %>%
  summarise(mean_Hawkishness = mean(hawkNumeric_player1), mean_benefit = sum(player_1_benefit, na.rm = TRUE))

perIndiv_player2 <- hdg %>%
  mutate(hawkNumeric_player2 = ifelse(player_2_card == "Hawk", 1, 0)) %>%
  group_by(teamID) %>%
  summarise(mean_Hawkishness = mean(hawkNumeric_player2), mean_benefit = sum(player_2_benefit, na.rm = TRUE))

perIndiv_bothPlayers <- bind_rows(perIndiv_player2,perIndiv_player1)
```



```{r hawkdoveresults2, echo = FALSE, results = 'hide'}
mod1 <- lm(meanHawk ~ round, data = hawkishness)
summary(mod1)
round(summary(mod1)$coefficients[2,4],3)
```

The average hawkishness overall is `r round(mean(hawkishness$meanHawk),3)`, but there is some variation among individuals.

```{r hawkdoveresults3, echo = FALSE, message = FALSE}
ggplot(perIndiv_bothPlayers,aes(x = mean_Hawkishness)) + 
  geom_histogram(bins = 13) + 
  xlim(0,1) + 
  ggtitle("There is variation in strategy among individuals") +
  xlab("Mean hawkishness")
```




We can now ask whether the level of hawkishness changed during the game on average across all the participants.

We can see from this graph that as the game proceeds round-by-round, there is no significant change in the level of hawkishness (p=`r round(summary(mod1)$coefficients[2,4],3)`). 

In other cases, it has been found that the average level of hawkishness tends to decline with the amount of time playing together, as opponents build trust and begin to see the value in cooperation.

```{r hawkdoveresults5, echo = FALSE, message = FALSE}
ggplot(hawkishness,aes(x = round,y = meanHawk)) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  ylim(0,1) + 
  ylab("Mean hawkishness") + 
  xlab("Round")
```
Next, we can ask whether the amount of benefits received (fitness) is associated with the degree of hawkishness. This, in evolutionary terms, is a way of seeing what the Evolutionary Stable Strategy (ESS) is. Is it an intermediate strategy, or more hawkish or dovish?

What we see is that the maximum benefits are received when hawkishness is at an intermediate level of around 0.5-0.6, though we clearly suffered from a lack of dove-like individuals in the study group!


```{r hawkdoveresults6, echo = FALSE, message = FALSE}
ggplot(perIndiv_bothPlayers,aes(x = mean_Hawkishness,y = mean_benefit)) + 
  geom_point() + 
  stat_smooth(method="lm",formula=y~poly(x, 2),fullrange=TRUE)+
  ggtitle("The best strategy in Game 1\nis an intermediate level of hawkishness") + 
  xlim(0,1) +
  ylab("Mean benefit") + 
  xlab("Mean hawkishness")
```

### Game 2 - different opponents

We didn't do this one in class this year, so here is what we found last year.

The students were asked to compete against a new opponent every round. The idea was that this would make it much harder to learn your opponents strategy and may be harder to come to a agreement that reduces aggression.


```{r hawkdoveresults7, echo = FALSE, warning = FALSE}
hdg2_candidates <- list.files(
  "data",
  pattern = "^gametheoryResults.*\\.xlsx$",
  full.names = TRUE
)

hdg2_path <- hdg2_candidates[which.max(file.info(hdg2_candidates)$mtime)]

hdg2 <- readxl::read_xlsx(hdg2_path, sheet = 4) %>%
  clean_names() %>%
  select(individual, round, player_1, player_2, benefit)
```




```{r hawkdoveresults8, echo = FALSE, results = 'hide'}
# Does hawkishness change during the game?
hawkishness <- hdg2 %>%
  mutate(hawkNumeric_player1 = ifelse(player_1 == "H", 1, 0)) %>%
  group_by(round) %>%
  summarise(meanHawk = mean(hawkNumeric_player1))

mod1 <- lm(meanHawk ~ round, data = hawkishness)
summary(mod1)
round(summary(mod1)$coefficients[2,4],3)


```

Firstly I have calculated the average hawkishness as  `r round(mean(hawkishness$meanHawk),3)`, which is not so different from the previous game.

Again, we can ask whether hawkishness changed during the game.

In this case, it DOES look like hawkishness declines more during the game. However, this trend is not significantly different from horizontal (p = `r round(summary(mod1)$coefficients[2,4],3)`).

```{r hawkdoveresults9, echo = FALSE}
ggplot(hawkishness,aes(x = round,y = meanHawk)) + 
  geom_point() + 
  geom_smooth(method = "lm")
```


Finally we can look at what the best strategy is in Game 2.



```{r hawkdoveresults10, echo = FALSE}
# How does the hawkishness of individuals relate to the benefits received?
perIndiv <- hdg2 %>%
  mutate(hawkNumeric_player1 = ifelse(player_1 == "H", 1, 0)) %>%
  group_by(individual) %>%
  summarise(meanHawkishness = mean(hawkNumeric_player1), benefit = sum(benefit, na.rm = TRUE))

ggplot(perIndiv,aes(x = meanHawkishness,y = benefit)) + 
  geom_point() + 
  geom_smooth(method="lm", formula=y~poly(x, 2))

```

It looks like the best strategy has changed a lot! Now the best strategy is to be a hawk.

In this setting we might expect  natural selection to drive towards the evolution of more aggressive individuals.

Think about what this means in terms of human cooperation and conflict avoidance. 

Can you see the value of understanding your opponent/competitor?
